The brightness and spatial distributions of terrestrial radio sources

### 2013-04-05 articles/Article-RFI-distribution/article-rfi-distribution.tex

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time-integration of radio observations if they are consistently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the integrated strength of RFI sources are consistent, i.e., they are observed as coherent sources. This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible after long-time integration. Consequently, it is likely that RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources. However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.

@@

\begin{figure*}
\begin{center}
\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
\caption{A small part of an observation displayed in a time-frequency plot. The features with significantly higher values are caused by interference. Some of these have a constant frequency, while others are more erratic.}
\label{fig:dist-rfi-example}
\end{center}
\end{figure*}

@@

\subsection{Calibration} \label{sec:calibration}
We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station. The current LOFAR SEFD is found to be approximately 3000~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz. For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 30,000~Jy. The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
\begin{equation}
 \sigma = \frac{\textrm{SEFD}}{\sqrt{2 \Delta \nu \Delta t}},
\end{equation}
where $\Delta \nu$ is the bandwidth and $\Delta t$ is the correlator integration time. The standard deviation will appear as the mode of the Rayleigh distribution. By fitting a Rayleigh function with fitting parameter $\sigma$ to the distribution, one finds the corresponding flux density scale.

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. We refer to van Haarlem et al. (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
% TODO correct reference if van Haarlem et al is published
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

\begin{table}
\caption{Data set specifications}\label{table:dist-data-specs}
\begin{center}
\begin{tabular}{lrr}
                    & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
Observation date    & 2011-10-09      & 2010-12-27 \\
Start time          & 06:50 UTC       & 0:00 UTC \\
Length              & 24 h           & 24 h \\
Time resolution     & 1 s             & 1 s \\
\hline
Frequency range     &  30.1--77.5 MHz & 115.0--163.3 MHz\\
Frequency resolution & 0.76 kHz    & 0.76 kHz \\
Number of stations  &  33           & 13 \\
Total size          & 96.3 TB        & 18.6 TB \\
\hline
Field               & NCP             & NCP \\
Amount RFI detected & 1.77\%       & 3.18\% \\
\hline
\hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

@@

@@

\subsection{Implications for very long integrations}
In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations. As an example, we will analyse the situation for the LOFAR EoR project. This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields. The final resolution required for signal extraction will be about $1$~MHz. The project will use about $60$~stations, each of which provides two polarized feeds. This will bring the noise level in a single 6~h observation in 1~MHz bandwidth to
\begin{equation}
 \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy}.
\end{equation}
Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.

@@

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.

>>>>

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long integrations of radio observations if they are consistently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the strength of RFI sources remain strong after integration, i.e., they are observed as coherent sources. This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible after long-time integration. Consequently, it is likely that RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

In time-frequency space, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a dynamic spectrum. Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in time-frequency space. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources. However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.

@@

\begin{figure*}
\begin{center}
\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
\caption{A small part of an observation displayed in a dynamic spectrum. The features with significantly higher values are caused by interference. Some of these have a constant frequency, while others are more erratic.}
\label{fig:dist-rfi-example}
\end{center}
\end{figure*}

@@

\subsection{Calibration} \label{sec:calibration}
We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station. The current LOFAR SEFD is found to be approximately 3400~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz. For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 34,000~Jy. The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
\begin{equation}
 \sigma = \frac{\textrm{SEFD}}{\sqrt{2 \Delta \nu \Delta t}},
\end{equation}
where $\Delta \nu$ is the bandwidth and $\Delta t$ is the correlator integration time. The standard deviation will appear as the mode of the Rayleigh distribution. By fitting a Rayleigh function with fitting parameter $\sigma$ to the distribution, one finds the corresponding flux density scale.

@@

\section{Data description} \label{sec:dist-data}
% TODO correct reference if van Haarlem et al is published
We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. We refer to van Haarlem et al. (2013, \textit{submitted}) for a full description of the capabilities of LOFAR. The analyses will cover only Dutch stations. Each Dutch station consists of 96 dipole low-band antennas (LBA) and one or two fields totalling 48~tiles of 4x4 bow-tie high-band antennas (HBA). The core area of LOFAR is located near the village of Exloo in the Netherlands, where the station density is at its highest. The six most densely packed stations are on the Superterp, an elevated area surrounded by water situated 3~km North of Exloo. A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.

In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities. In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

\begin{table}
\caption{Data set specifications}\label{table:dist-data-specs}
\begin{center}
\begin{tabular}{lrr}
                    & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
Observation date    & 2011-10-09      & 2010-12-27 \\
Start time          & 06:50 UTC       & 0:00 UTC \\
Length              & 24 h           & 24 h \\
Time resolution     & 1 s             & 1 s \\
\hline
Frequency range     &  30.1--77.5 MHz & 115.0--163.3 MHz\\
Frequency resolution & 0.76 kHz    & 0.76 kHz \\
Number of stations  &  33           & 13 \\
Total size          & 96.3 TB        & 18.6 TB \\
\hline
Field               & NCP             & NCP \\
Amount of RFI detected & & \\
by the AOFlagger    & 1.77\%       & 3.18\% \\
\hline
\hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2,700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\multicolumn{4}{l}{\textit{Average station temperatures}} \\
\hline
$T_\textrm{sys}$ & System temperature (in clean bands)  & 5,000 K & 640 K \\
$T_\textrm{R}$ & Temperature of RFI & 17,000 K & 1,200 K \\
$T_\textrm{leak}$ & Temperature of undetected RFI & 3.2 K & 1.4 K \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79~m high transmitters nearby LOFAR stations in the LBA frequency range. Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law. This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively. Therefore, the LBA measurements are clearly not accurate enough to be conclusive. Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.

@@

\citet{lf-interference-temperature-rogers} TODO

@@

\subsection{Implications for very long integrations}
In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations. As an example, we will analyse the situation for the LOFAR EoR project. This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields. The final resolution required for signal extraction will be about $1$~MHz. The project will use about $60$~stations, each of which provides two polarized feeds. This will bring the noise level in a single 6~h observation in 1~MHz bandwidth to
\begin{equation}
 \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy},
\end{equation}
where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers. Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.

@@

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
